G\"{o}del's God on the Computer (Abstract)\thanks{This work has been supported by the German Research Foundation under grant BE2501/9-1.}

### 2013-11-04 Abstract/IWIL-2013/abstract.tex
more press links
update of IWIL abstract
several updates wrt to tpi script and Isabelle session

\authorrunning{C.\~Benzm\"{u}ller \and B.\~Woltzenlogel Paleo}
\titlerunning{G\"{o}del's God on the Computer}

@@

G\"{o}del's proof is challenging to formalize and verify because it
requires an expressive logical language with modal operators
(\emph{possibly} and \emph{necessarily}) and with
quantifiers for individuals and properties.  Our computer-assisted formalizations rely on an
embedding of the modal logic into classical higher-order logic with
Henkin semantics. The formalization is thus essentially
done in classical higher-order logic where  quantified modal logic is
emulated.

In our ongoing computer-assisted study of G\"odel's proof, the
automated reasoning tools have made some interesting observations,
some of which were unknown so far. This attests the maturity of
contemporary interactive and automated deduction tools for classical
higher-order logic and demonstrates the elegance and practical
relevance of the embeddings-based approach.  Most importantly, our
work opens new perspectives for a computer-assisted theoretical
philosophy.  The critical discussion of the underlying concepts,
definitions and axioms remains a human responsibility, but the
computer can assist in building and checking rigorously correct
logical arguments. In case of logico-philosophical disputes, the
computer can check the disputing arguments and partially fulfill
Leibniz' dictum: Calculemus --- Let us calculate!

\small

@@

\bibitem{AndersonGettings}
A.C. Anderson and M.~Gettings.
\newblock G\"odel ontological proof revisited.
\newblock In {\em {G\"odel'96: Logical Foundations of Mathematics, Computer
  Science, and Physics: Lecture Notes in Logic 6}}. {Springer}, 1996.

\bibitem{B9}
C.~Benzm{\"u}ller and L.C. Paulson.
\newblock Exploring properties of normal multimodal logics in simple type
  theory with {LEO-II}.
\newblock In {\em {Festschrift in Honor of {Peter B. Andrews} on His 70th
  Birthday}}, pages 386--406. College Publications.

@@

\bibitem{LEO}
C.~Benzm{\"u}ller, F.~Theiss, L.~Paulson, and A.~Fietzke.
\newblock {LEO-II} - a cooperative automatic theorem prover for higher-order
  logic.
\newblock In {\em Proc. of IJCAR 2008}, volume 5195 of {\em LNAI}, pages
  162--170. Springer, 2008.

@@

\bibitem{Satallax}
C.E. Brown.
\newblock Satallax: An automated higher-order prover.
\newblock In {\em Proc. of IJCAR 2012}, number 7364 in LNAI, pages 111 -- 117.
  Springer, 2012.

@@

% \bibitem{Hazen}
% A.P. Hazen.
% \newblock On g\"odel's ontological proof.
% \newblock {\em Australasian Journal of Philosophy}, 76:361--377, 1998.

\bibitem{Metis}
J.~Hurd.
\newblock First-order proof tactics in higher-order logic theorem provers.
\newblock In {\em Design and Application of Strategies/Tactics in Higher Order
  Logics, NASA Tech. Rep. NASA/CP-2003-212448}, 2003.

@@

% \bibitem{FormalTheologyRepository}
% B.~Woltzenlogel Paleo and C.~Benzm\"uller.
% \newblock Formal theology repository
%   (\url{https://github.com/FormalTheology/GoedelGod}).

@@

@@

\bibitem{TPI}
G.~Sutcliffe.
\newblock The TPTP TPI Scripting Language (check!).
\newblock {\em Proc. of IWIL-10}, 2013.

>>>>

\authorrunning{C. Benzm\"{u}ller and B. Woltzenlogel Paleo}
\titlerunning{G\"{o}del's God on the Computer}

@@

G\"{o}del's proof is challenging to formalize and verify because it
requires an expressive logical language with modal operators
(\emph{possibly} and \emph{necessarily}) and with
quantifiers for individuals and properties.  Our computer-assisted formalizations rely on an
embedding of the modal logic into classical higher-order logic with
Henkin semantics  \cite{J23,B9}. The formalization is thus essentially
done in classical higher-order logic where  quantified modal logic is
emulated.

In our ongoing computer-assisted study of G\"odel's proof, the
automated reasoning tools have made some interesting observations,
some of which were unknown so far. Our work attests the maturity of
contemporary interactive and automated deduction tools for classical
higher-order logic and demonstrates the elegance and practical
relevance of the embeddings-based approach.  Most importantly, our
work opens new perspectives for a computer-assisted theoretical
philosophy.  The critical discussion of the underlying concepts,
definitions and axioms remains a human responsibility, but the
computer can assist in building and checking rigorously correct
logical arguments. In case of logico-philosophical disputes, the
computer can check the disputing arguments and partially fulfill
Leibniz' dictum: Calculemus --- Let us calculate!

@@

\bibitem{AndersonGettings}
A.C. Anderson and M.~Gettings.
\newblock G\"odel ontological proof revisited.
\newblock In {\em {G\"odel'96: Logical Foundations of Mathematics, Computer
  Science, and Physics: Lecture Notes in Logic 6}}, pages 167--172. {Springer},
  1996.

\bibitem{B9}
C.~Benzm{\"u}ller and L.C. Paulson.
\newblock Exploring properties of normal multimodal logics in simple type
  theory with {LEO-II}.
\newblock In {\em {Festschrift in Honor of {Peter B. Andrews} on His 70th
  Birthday}}, pages 386--406. College Publications, 2008.

@@

\bibitem{LEO}
C.~Benzm{\"u}ller, F.~Theiss, L.C. Paulson, and A.~Fietzke.
\newblock {LEO-II} - a cooperative automatic theorem prover for higher-order
  logic.
\newblock In A.~Armando, P.~Baumgartner, and G.~Dowek, editors, {\em Automated
  Reasoning, 4th International Joint Conference, IJCAR 2008, Sydney, Australia,
  August 12-15, 2008, Proceedings}, volume 5195 of {\em LNCS}, pages 162--170.
  Springer, 2008.

@@

\bibitem{Satallax}
C.E. Brown.
\newblock Satallax: An automatic higher-order prover.
\newblock In B.~Gramlich, D.~Miller, and U.~Sattler, editors, {\em Automated
  Reasoning, 6th International Joint Conference, IJCAR 2012, Manchester, UK,
  June 26-29, 2012. Proceedings}, volume 7364 of {\em LNCS}, pages 111--117.
  Springer, 2012.

@@

\bibitem{Metis}
J.~Hurd.
\newblock First-order proof tactics in higher-order logic theorem provers.
\newblock In {\em Design and Application of Strategies/Tactics in Higher Order
  Logics, NASA Tech. Rep. NASA/CP-2003-212448}, pages 56--68, 2003.

@@

@@

\bibitem{TPI}
G.~Sutcliffe.
\newblock {The TPTP Process Instruction Language}.
\newblock In S.~Schulz, G.~Sutcliffe, and B.~Konev, editors, {\em {Proceedings
  of the 10th International Workshop on the Implementation of Logics}},
  EasyChair Proceedings in Computing, 2013.
\newblock To appear.

@@
